\documentclass[12pt,a4paper,pdftex]{article}

\usepackage[top=1in, bottom=1in, left=1.4in, right=1.2in]{geometry}

% Paragraphs use a vertical space rather than tabbing
\setlength{\parskip}{\medskipamount}
\setlength{\parindent}{0pt}
\usepackage[hang,flushmargin]{footmisc} % same for footnotes

% Title within the first page
\usepackage{titling}   %Reduce spacing before title
\setlength{\droptitle}{-6em}   %-6em reduces to ~normal margins

\usepackage[colorlinks=true, % colour all links black (just using
linkcolor=black, % colorlinks=false results in boxes around
citecolor=black, % links)
filecolor=black,
urlcolor=black,
breaklinks]{hyperref}


\usepackage{xr}
\externaldocument{main}


% TC:ignore
\usepackage[sort&compress, % on multiple refs sort them and write as range
capitalise, % Use Section not section etc.
noabbrev, % Use Table not Tab. etc.
nameinlink % Make the name (eg Section) part of the hyperlink
]{cleveref}
% TC:endignore

% Call subsections sections
\crefname{subsection}{Section}{Sections}
\Crefname{subsection}{Section}{Sections}

% just use (...) for equations
\crefformat{equation}{#2(#1)#3}
\crefrangeformat{equation}{#3(#1)#4--#5(#2)#6}
\crefmultiformat{equation}{(#2#1#3)}{ and~(#2#1#3)}{, (#2#1#3)}{ and~(#2#1#3)}

% Except for start of sentences where we need to say "Equations"
\Crefformat{equation}{Equation~#2(#1)#3}
\Crefrangeformat{equation}{Equations~#3(#1)#4--#5(#2)#6}
\Crefmultiformat{equation}{Equations~(#2#1#3)}{ and~(#2#1#3)}{, (#2#1#3)}{ and~(#2#1#3)}

% a reference for "this X"
\newcommand{\thisref}[1]{this \lcnamecref{#1}}
\newcommand{\Thisref}[1]{This \lcnamecref{#1}}


\input{latex-commands/commands.tex}


\title{Thesis Corrections}
\author{David Shepherd}

\begin{document}
\maketitle


\begin{enumerate}
\item
  \begin{quotation}
    Implement minor corrections (typos etc) as provided through
    separate list or marked up in thesis from both examiners.
  \end{quotation}
  I have implemented the minor corrections.

\item
  \begin{quotation}
    Chapter 7: Extend the list of methods tested to include one easily
    available library functions, such as scipy's VODE (both for 7 and
    8.1). Given the many libraries that are easily available, the
    practitioner will ask whether it is worth using (and implementing) the
    IMR method. Thus the direct comparison will embed the new IMR results
    into the research context. Furthermore, comparison of the methods
    coded by yourself with established library code will instill further
    trust in the correctness of your coding and the results.
  \end{quotation}
  I have compared scipy's \vode BDF2 integrator against \oomph's BDF2 integrator for an ODE problem and, as would be expected, the new results are almost identical. The results and some discussion are included in the chapter on the adaptive implicit midpoint rule (on page \pageref{fig:vode-osc-example}).

  It should also be noted that \oomph's BDF2 integrator was not implemented by me, in fact it is the main time integration scheme of \oomph.
  It has been thoroughly tested over the past $\sim 10$ years and used has been used in numerous papers and theses.

\item
  \begin{quotation}
    Introduce methods [renormalization to orbit of precession],
    [self-correcting LLG] and [tolerance based renormalisation], somewhere
    in the thesis (for example section 3.4.8) as practical "state of the
    art" approaches that have been proposed in the community to ensure
    magnetisation length conservation.
  \end{quotation}
  The self-correcting LLG and a tolerance-based renormalisation approach are now introduced in section \ref{sec:ensuring-constant-mv} (Magnetisation length conservation) which begins on page \pageref{alt-ml-renorm-intro}.

  On the subject of the ``normalisation to the orbit of precession'' method I have not been able to find any information in the literature beyond the bullet points in the presentation.
  I have asked Mike Donahue for more information but he was unable to provide any.
  As such I have not been able to introduce this method.

\item
  \begin{quotation}
    Review section 7.4 and 8, using the above re-normalisation techniques
    to provide better context for the IMR method. In more detail: repeat
    the computational examples with [the methods] and compare with the IMR.
    It seems sensible to keep the current data on (i) no renormalisation
    and (ii) re-normalisation after every step to provide a comprehensive
    overview of the different possible approaches.
  \end{quotation}
  I have carried out detailed experiments comparing the tolerance based renormalisation and self-correcting LLG methods with the basic method of re-normalisation after every step for the example ODE LLG problem.
  These experiments are documented in a new 9 page appendix (appendix \ref{sec:magn-renorm-meth} on page \pageref{sec:magn-renorm-meth}).
  The conclusion is that the basic re-normalisation method is clearly superior to the alternative methods, and there is no reason to expect that these results would change for other problems.
  As such there does not seem to be any reason to experiment with the alternative methods for other example problems.

% ??ds first draft
\item
  \begin{quotation}
    Consider a review of the choice of figures in chapter 8 to convey
    the intended message(s) as clearly as possible; i.e. choose figures
    that explain the value of the IMR method. Measures such as CPU time
    used overall (or Wall clock time) and memory usage are particularly
    attractive as these are key factors in day-to-day application of
    micromagnetic modelling, in addition to measures to summarise the
    accuracy of the methods. Numbers of required iterations etc are
    interesting to better understand where a benefit/disadvantage comes
    from.
  \end{quotation}
  Since the value of the IMR method is likely to come from a reduction in the global (temporal) error norm I have added additional passages to chapters 7 and 8 (on pages \pageref{cpu-correction-1}, \pageref{cpu-correction-2}) detailing how differences in this error norm of the time integration method affect the CPU time via the ability to take larger time steps.

It would be extremely time consuming to create figures showing this reduction in CPU time directly: in order to fairly compare the CPU time used by a time integration scheme one would have to somehow ensure that the global error norms for the results were almost identical.
Essentially this could only be done by trial and error, and due to the non-linearity of the problem it may be impossible.

  Plots of the numbers of iterations are already shown for the example problems where novel solvers are used.
However, I have added descriptions of how the number of Newton-Raphson iterations and linear solver iterations affect the memory (via reduced maximum Krylov subspace size) and CPU time on page \pageref{mem-cpu-solver-correction}.

% ??ds first draft
\item
  \begin{quotation}
    In the mumag standard problem 4 study, we suggest not to compare
    against d'Aquino's data as this is based on finite difference
    calculations and very likely to introduce different dynamical behaviour
    due to the different computation of demag and the other fields.
    Instead, repeat the simulation with the different integrator
    techniques, i.e IMR, and [the three alternative renormalisation
    methods], where [each renormalisation method] can each be combined with
    TR, BDF2 and - if desired - the CVODE approach. This will allow you to
    look at the changes in great detail and any difference is guaranteed to
    originate from the integration routine (as you use the same
    computational framework for everything else).
  \end{quotation}
  I have added additional results for the \mumag problem \#4 solved with alternative time integration methods as requested.
  A straightforward comparison of TR, BDF2 and IMR in terms of the magnetisation values and the time steps selected has been added on page \pageref{more-mumag4-correction-1}.
  On page \pageref{more-mumag4-correction-2} I have added some results comparing the accuracy of the nodal and Gaussian quadrature schemes (this is very relevant to the ``comparison of different integrator techniques'' which was requested because geometric integration requires both IMR and a nodal quadrature scheme).

  However, while I agree that the comparison with finite difference methods is potentially harsher than necessary, I still believe that it is important to verify the model by comparing against other models for some standard problem.
  Since the \mumag problem \#4 is the only such standard problem for dynamic micromagnetics, and since all available data for the solution are computed using finite difference methods there is no better choice for this comparison.
  As such I would strongly prefer to leave the comparison in my thesis.

\item
  \begin{quotation}
    Rewrite the discussion of the energy conservation error for the
    triangular elements to draw out the positives. Although the
    conservation does not appear to be as good as quad elements, it is
    still better than the current state-of-the-art using standard time
    integration methods.
  \end{quotation}
  I have rewritten the discussion of energy conservation with triangular elements in a more positive manner as requested (see pages \pageref{sec:triangular-meshes}, \pageref{sec:numer-experiments-conclusions}, \pageref{sec:future-work}).

\end{enumerate}

% Were the plots in chapter 6 included as a correction or just a suggestion for the future? They don't appear in either of the typed lists of corrections and there is only a question on Andrew's handwritten copy?
% The possible correction was to replace plots by means + stddev. When you
% do this remember to also change the plots in ch. 8

\end{document}


%%% Local Variables:
%%% mode: latex
%%% TeX-master: ""
%%% End:
