\documentclass[12pt,PhD,a4paper,twoside,pdftex]{muthesis}

% Define a boolean which says we are compiling my thesis
\newif\ifthesis
\thesistrue


%% General packages
%% ============================================================

% Add maths commands, symbols and fonts
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsfonts}
\usepackage{amsthm}
\usepackage{mathtools}

% Greek letters in text
\usepackage[euler]{textgreek}


\usepackage[linesnumbered]{algorithm2e}


\usepackage{url}                %Support for urls

% Allow hyperlinks to break over lines and turn off warning about it
\usepackage{silence}
\WarningFilter{hyperref}{You have enabled option `breaklinks'.}
\usepackage[hyphenbreaks]{breakurl}

% This stops the word count from complaining that it doesn't recognise
% packages. Note that the lack of a space in "%TC" is important!
%TC:ignore
\usepackage[colorlinks=true, % colour all links black (just using
            linkcolor=black, % colorlinks=false results in boxes around
            citecolor=black, % links)
            filecolor=black,
            urlcolor=black,
            breaklinks]{hyperref}
%TC:endignore

\usepackage{xspace} % command for fixing spaces after macros




% Language
% ============================================================
% Support for UTF8 input
\usepackage[utf8]{inputenc}

%% % Handle non-ascii fonts properly, see
%% % http://tex.stackexchange.com/questions/664/why-should-i-use-usepackaget1fontenc
%% \usepackage[T1]{fontenc}
%% % Seems to break searching so I'm not using it...

% Commonly recommend, not sure what it does...
\usepackage[UKenglish]{babel}



%% Page Layout
%% ============================================================
% Paragraphs use a vertical space rather than tabbing
\setlength{\parskip}{\medskipamount}
\setlength{\parindent}{0pt}
\usepackage[hang,flushmargin]{footmisc} % same for footnotes

% Adds a landscape enviroment
\usepackage{lscape}


%% Figure Tweaks
%% ============================================================
\usepackage[pdftex]{graphicx}   %Needed to include pictures

% positioning tweaks
\renewcommand{\topfraction}{0.85}
\renewcommand{\textfraction}{0.1}
\renewcommand{\floatpagefraction}{0.75}

% Italic captions
\usepackage[labelfont=it,textfont=it]{caption}

% package for \FloatBarrier which stops floats moving past it.
\usepackage{placeins}



%% Equations/maths
%% ============================================================
\numberwithin{equation}{chapter} % Number equations by "<chapter>.<number>".

%\usepackage{nath}
% % Prevent line break in the middle of maths except in extreme cases (increse numbers to make line breaking even more unlikely)
% \relpenalty=3000
% \binoppenalty=5000



%% Bibliograpy Tweaks
%% ============================================================

% biblatex setup
%TC:ignore
\usepackage[backend=biber,
  citestyle=numeric,
  firstinits=true, % Only print initials of first names
  isbn=false,
  doi=true,
  url=true, % for websites, see below for removing it from papers etc.
  % backref=true
]{biblatex}
%TC:endignore


% clear urls for papers, proceedings, books
\AtEveryBibitem{%
  \ifentrytype{article}{%
    \clearfield{url}%
    \clearfield{urldate}%
  }
  {}% no "else" operation
  %
  \ifentrytype{inproceedings}{%
    \clearfield{url}%
    \clearfield{urldate}%
  }
  {}% no "else" operation
  %
  \ifentrytype{book}{%
    \clearfield{url}%
    \clearfield{urldate}%
  }
  {}% no "else" operation
}

% Get rid of "In:" for journal articles
\renewbibmacro{in:}{%
  \ifentrytype{article}{}{\printtext{\bibstring{in}\intitlepunct}}}



% Add bibtex files
\addbibresource{mendeley_library.bib}

%% % A superscript ciation command
%% \DeclareCiteCommand{\supercite}[\mkbibsuperscript]
%%   {\iffieldundef{prenote}
%%      {}
%%      {\BibliographyWarning{Ignoring prenote argument}}%
%%    \iffieldundef{postnote}
%%      {}
%%      {\BibliographyWarning{Ignoring postnote argument}}}
%%   {\usebibmacro{citeindex}%
%%    \bibopenbracket\usebibmacro{cite}\bibclosebracket}
%%   {\supercitedelim}
%%   {}

%% % Use superscript cite instead of normal ones
%% \let\cite=\supercite



% Section etc. labelling
% ============================================================

% This goes last because apparently other packages can break cleveref

% \def\chapterautorefname{Chapter} % capitalise
% \def\sectionautorefname{Section} % capitalise
% \def\subsectionautorefname{Section} % capitalise, say section not subsection

%TC:ignore
\usepackage[sort&compress, % on multiple refs sort them and write as range
            capitalise, % Use Section not section etc.
            noabbrev, % Use Table not Tab. etc.
            nameinlink % Make the name (eg Section) part of the hyperlink
            ]{cleveref}

%TC:endignore

% Call subsections sections
\crefname{subsection}{Section}{Sections}
\Crefname{subsection}{Section}{Sections}

% just use (...) for equations
\crefformat{equation}{#2(#1)#3}
\crefrangeformat{equation}{#3(#1)#4--#5(#2)#6}
\crefmultiformat{equation}{(#2#1#3)}{ and~(#2#1#3)}{, (#2#1#3)}{ and~(#2#1#3)}

% Except for start of sentences where we need to say "Equations"
\Crefformat{equation}{Equation~#2(#1)#3}
\Crefrangeformat{equation}{Equations~#3(#1)#4--#5(#2)#6}
\Crefmultiformat{equation}{Equations~(#2#1#3)}{ and~(#2#1#3)}{, (#2#1#3)}{ and~(#2#1#3)}


% a reference for "this X"
\newcommand{\thisref}[1]{this \lcnamecref{#1}}
\newcommand{\Thisref}[1]{This \lcnamecref{#1}}



\setcounter{tocdepth}{1} % Set contents to only go down to section level


%% Flow charts
%% ============================================================
\usepackage{tikz} % Package for drawing flow charts
\usetikzlibrary{shapes,arrows}

\definecolor{paleblue}{RGB}{239,242,255}
\definecolor{solidblue}{RGB}{43,0,229}

% Define a rectangle shape and an arrow
\tikzstyle{block} = [rectangle, draw, fill=paleblue, thick,
    text width=4.3cm, text centered, rounded corners, minimum height=1cm]
\tikzstyle{line} = [draw, -]
\tikzstyle{arrow} = [draw, -latex']



% Define my commands
% ============================================================
\input{commands.tex}


%% Beginning of main document
%% =============================================================

% Include only some sections while working on them, remove before final
%\includeonly{first_year_progress}

\title{Numerical methods for micromagnetics}
\author{David Shepherd}
\principaladviser{Jim Miles\\
  Milan Mihajlovi\'{c}\\
  Matthias Heil}
% \submitdate{29th September 2014}

% \date{30 March 2012}


\begin{document}

% muthesis stuff:
%%%%%%%%%%

% disable some useless pages
% \figurespagefalse
\tablespagefalse
% \copyrightfalse

\beforeabstract
\abstracttitle
Micromagnetics is a continuum mechanics theory of magnetic materials widely used in industry and academia.
In this thesis we describe a complete numerical method, with a number of novel components, for the computational solution of dynamic micromagnetic problems by solving the Landau-Lifshitz-Gilbert  (LLG) equation.
In particular we focus on the use of the implicit midpoint rule (IMR), a time integration scheme which when applied to the LLG conserves magnetisation length and, in the case of zero damping, the energy.
We use the finite element method (FEM) for spatial discretisation, and use nodal quadrature schemes to retain the conservation properties of IMR despite the weak-form approach.

We introduce a novel, generally applicable adaptive time step selection algorithm for the IMR.
We show that our scheme selects error-appropriate time steps for a variety of problems and that it retains the conservation of magnetisation length and conservation of energy properties of the fixed step IMR.
We also show how these conservation properties can be extended to the PDE case using finite elements with a nodal quadrature scheme.

We demonstrate how hybrid FEM/BEM magnetostatic calculations can be coupled to the LLG equation in a monolithic manner.
This allows the coupled solver to maintain all properties of the standard time integration scheme, in particular the energy conservation property of IMR and the solution converged to in the stochastic case.
We also develop a preconditioned Krylov solver for the coupled system which can efficiently solve the monolithic system given an effective preconditioner for the decoupled LLG system.

Finally we investigate the effect of the spatial discretisation on the comparative effectiveness of implicit and explicit time integration schemes (\ie the stiffness).
We find that explicit methods are more efficient for simple problems, but for the fine spatial discretisations, required in a number of more complex cases, implicit schemes become much more efficient.

\afterabstract

\prefacesection{Acknowledgements}
I would like to thank...

supervisors

Ray

Sylwia

family

NEST group

\oomph group

and all the people I met....

epsrc funding

\afterpreface

%%%%%%%%%% % end of muthesis stuff


\include{introduction}
\include{continuous_micromagnetics}
\include{intro-numerical-methods}
\include{fem-llg}
\include{boundary-elements}

\include{fem-bem-linear-systems}
\include{adaptive-imr}
\include{numerical-experiments}
\include{llg-stiffness-analysis}

\include{conclusions-future-work}

\appendix
\include{analytical-solutions}
\include{h-linear-sym-operator}
\include{truncation-errors}



% Does what it says (using biblatex package)
\printbibliography

\end{document}


%%% Local Variables:
%%% mode: latex
%%% TeX-master: t
%%% End:
