
\section{Non-dimensionalisation}
\label{sec:normalisations-appendix}

%??ds talk about why?

% no unecessary (for the maths) parameters
% improves accuracy by minimising round off error

\subsection{The Landau-Lifshitz-Gilbert equation}
\label{sec:land-lifsh-gilb-normalisation}

We start from the Landau--Lifshitz--Gilbert equation with the magnetostatic, applied, exchange and magnetocrystalline (effective) fields (see Section~\ref{sec:cont-micromag}):

\begin{align}
  \pd{\Mv^*}{t^*} &= - \gymagc \Mv^* \times \Hv^* + \frac{\alpha}{M_s} \Mv^* \times \pd{\Mv^*}{t^*}, \label{eqn:llgnd} \\
  \Hv^* &= \Happ^* - \nabla^* \phi^* + \frac{2\Exchc^*}{\mu_0 M_s} \nabla^{*2} \mv + \frac{2\Kone^*}{\mu_0 M_s} (\mv \cdot \ev) \ev.
  \label{eqn:effnd} \\
  \nabla^{*2} \phi^* &= \nabla^* \cdot \Mv^* \label{eqn:phind}
\end{align}
where we use $^*$ to denote the dimensional variables, operators and constants that will be non-dimensionalised. We assume (for simplicity) that $M_s$, $\Exchc$ and $\Kone$ are all
constant throughout the magnetic materials used.

Let
\begin{align}
  \Mv^* &= M_s \mv, \notag \\
  \phi^* &= \Phi \phi, \notag \\
  \Hv^* &= M_s \hv, \label{eqn:nddefs} \\
  \Exchc^* &= \nA \exchc, \notag \\
  \Kone^* &= \nK \kone, \notag \\
  t^* &= \frac{1}{\gymagc M_s} t, \notag \\
  x_i^* &= l x_i. \notag
\end{align}
Note that $\Mv$ and $\Hv$ have the same units so we use the same normalisation factor. For dimensional purposes derivatives are equivalent to division by the variable differentiated with respect to, so $\nabla^* = \frac{1}{l} \nabla$, $\pd{a}{t^*} = \gymagc M_s \pd{a}{t}$ etc.

Combining equation~\eqref{eqn:llgnd} with the definitions~\eqref{eqn:nddefs} gives
\begin{equation}
  \notag
  \dmdt \gymagc M_s^2 =
  - \gymagc M_s^2 (\mv \times \hv) + \frac{\alpha}{M_s} \gymagc M_s^3 (\mv \times \dmdt),
\end{equation}

cancelling the various constants results in the non-dimensionalised Landau--Lifshitz--Gilbert equation
\begin{equation}
  \label{eq:53}
  \dmdt = - (\mv \times \hv) + \alpha (\mv \times \dmdt).
\end{equation}

Similarly for equation~\eqref{eqn:phind}
\begin{align*}
  \frac{1}{l^2} \Phi \nabla^{2} \phi &= \frac{1}{l} M_s \nabla \cdot \mv, \\
  \frac{\Phi}{M_s l} \nabla^{2} \phi &= \nabla \cdot \mv.
\end{align*}

Letting $\Phi = M_s l$ we obtain
\begin{equation}
  \label{eq:57}
  \nabla^2 \phi = \nabla \cdot \mv.
\end{equation}

Repeating the substitutions for equation~\eqref{eqn:effnd} gives
\begin{align*}
  M_s \hv &= M_s \happ - \frac{\Phi}{l} \nabla \phi + \frac{2\nA}{\mu_0l^2 M_s}  \exchc \nabla^2 \mv + \frac{2\nK}{\mu_0 M_s}  \kone (\mv \cdot \ev) \ev, \\
  \hv &= \frac{M_s}{M_s} \happ - \frac{M_s}{M_s} \nabla \phi + \frac{2\nA}{\mu_0 l^2 M_s^2}  \exchc \nabla^2 \mv + \frac{2\nK}{\mu_0 M_s^2} \kone (\mv \cdot \ev) \ev.
\end{align*}
This can be further simplified by using:
\begin{align}
  \nK &= \mu_0 M_s^2, \notag \\
  \nA &= l^2 \nK, \label{eqn:ndfields}
\end{align}

leaving
\begin{equation}
  \hv = \happ - \nabla \phi + 2\exchc \nabla^2 \mv + 2\kone (\mv \cdot \ev) \ev.
\end{equation}

The parameter for length, $l$ is still free to be chosen. Useful possibilities are the exchange length or a unit on the length scale of a dimension of the magnetic material being modelled (typically $\sim1-100$nm).


\subsection{Energy calculations}
\label{sec:energy-calculations}

Sometimes we may need to know the non-dimensional energy of a micromagnetic state in a form consistent with that used for the LLG equation. One example is for the computation of an effective damping constant, see section ??ds.

We repeat the process used in Section~\ref{sec:land-lifsh-gilb-normalisation} to get a set of non-dimensionalised energy equations, starting from the equations given in Section~\ref{sec:energy-magnetic-body}:

\begin{equation*}
  \Eapp^* = - \mu_0 \int_{\magd} \Mv^* \cdot \Happ^* \d \magd^*,
\end{equation*}
\begin{equation}
  \Ems^* =  \frac{-\mu_0}{2} \int_{\magd} \Mv^* \cdot \Hms^* \d \magd^*,
\end{equation}
\begin{equation*}
  \Eex^* =  \Exchc^* \int_{\magd} (\nabla^* \mv)^2 \d \magd^*,
\end{equation*}
\begin{equation*}
  \Eca^* =  \Kone^* \int_\magd (\mv \cdot \ev)^2 \d \magd^*.
\end{equation*}

We substitute definitions~\eqref{eqn:nddefs}, \eqref{eqn:ndfields} and $E^* = \nE e = \mu_0 M_s^2 l^d \, e$ where $d$ denotes the number of spatial dimensions:
\begin{equation*}
  \eapp = \frac{-\mu_0 M_s^2 l^d}{\nE} \int_{\magd} \mv \cdot \happ \d \magd
  = - \int_{\magd} \mv \cdot \happ \d \magd,
\end{equation*}
\begin{equation}
  \ems = \frac{-\mu_0 M_s^2 l^d}{2\nE} \int_{\magd} \mv \cdot \hms \d \magd
  = - \int_{\magd} \mv \cdot \hms \d \magd,
\end{equation}
\begin{equation*}
  \eex =  \frac{\mu_0 M_s^2 l^2 l^d}{l^2\nE} \exchc \int_{\magd} (\nabla \mv)^2 \d \magd
  = \exchc \int_{\magd} (\nabla \mv)^2 \d \magd,
\end{equation*}
\begin{equation*}
  \eca = \frac{\mu_0 M_s^2 l^d}{\nE} \kone \int_\magd 1 - (\mv \cdot \ev)^2 \d \magd
  = \kone \int_\magd 1 - (\mv \cdot \ev)^2 \d \magd.
\end{equation*}
Note that we have used
\begin{equation}
  \d \magd^* = \d x^{*d} = l^d \d x^d = l^d \d \magd.
\end{equation}



%%% Local Variables:
%%% mode: latex
%%% TeX-master: "main"
%%% End:
