\documentclass[12pt,a4paper,pdftex]{article}

\usepackage[top=1in, bottom=1in, left=1.4in, right=1.2in]{geometry}

% Paragraphs use a vertical space rather than tabbing
\setlength{\parskip}{\medskipamount}
\setlength{\parindent}{0pt}
\usepackage[hang,flushmargin]{footmisc} % same for footnotes

% Title within the first page
\usepackage{titling}   %Reduce spacing before title
\setlength{\droptitle}{-6em}   %-6em reduces to ~normal margins


\input{../commands.tex}


\title{Corrections}
\author{David Shepherd}

\begin{document}
\maketitle

\begin{enumerate}
\item I have implemented the minor corrections.

\item I have compared scipy's \VODE BDF2 integrator against \oomph's BDF2 integrator for an ODE problem and, as would be expected, the results are almost identical. The results and some discussion are included in the chapter on adaptive implicit midpoint rule (page ??ds).

It should also be noted that \oomph's BDF2 integrator was not implemented by me.
It is the main time integration scheme of \oomph and has been thoroughly tested over the past $\sim 10$ years ??ds - Matthias how old is this code?.

\item The self-correcting LLG and a tolerance-based renormalisation approach are now introduced in the chapter standard numerical methods (page ??ds).

  On the subject of the normalisation to the orbit of precession I have not been able to find any information in the literature beyond the bullet points in the presentation.
I asked Mike Donahue for more information and he was unable to help ??ds - double check..
As such I have not yet been able to introduce this method.

\item I have carried out detailed experiments comparing the alternative normalisation methods with the basic method of re-normalisation after every step for the example ODE LLG problem.
These experiments are documented in a new 9 page appendix (page ??ds).
The conclusion is that the basic re-normalisation method is clearly superior to the alternative methods.
As such there does not seem to be any reason to experiment with PDE problems.

\item I have added additional passages to chapters 7 and 8 (pages ??cpu-correction) detailing how differences in the accuracy of the time integration method affect the CPU time via the ability to take larger time steps.
  Additionally in chapter 8 (page ??mem-cpu-solver-correction) I have described how the number of Newton-Raphson iterations and linear solver iterations affect the memory (via reduced maximum Krylov subspace size) and CPU time.

\item ??ds TODO
I agree that the comparison with finite difference methods is potentially misleading and so I have added a comparison with \nmag's results for the \mumag standard problem \#4.

??ds TODO
I have also added additional plots comparing with the results when using TR and BDF2.

\item I have rewritten discussion of energy conservation with triangular elements in a more positive manner, see page ??ds.

\end{enumerate}

Were the plots in chapter 6 included as a correction or just a suggestion for the future? They don't appear in either of the typed lists of corrections and there is only a question note on Andrew's handwritten copy?
% The possible correction was to replace plots by means + stddev. When you
% do this remember to also change the plots in ch. 8

\end{document}


%%% Local Variables:
%%% mode: latex
%%% TeX-master: ""
%%% End:
